\ripple: Differentiable \dfm{and Hardware-Accelerated} Waveforms for Gravitational Wave Data Analysis

### 2023-01-20 src/tex/ms.tex
Edits Section 3

Here, we will illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
In the paper we will primarily look at toy examples, leaving a more careful analyses to future work. 
The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to forecasting and parameter estimation.

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are commonly
regarded as the most accurate, they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
are only approximations but can be evaluated much faster.

Every waveform approximant has some fitting coefficients that are tuned during
the process of calibrating the ansatz to NR waveforms. In the case of
IMRPhenomD, there are 209 fitting coefficients used to capture the separate
behavior of the amplitude and phase as a function of the mass ratio and spins.
The accuracy of fitting coefficients determines how well the approximant
reproduces the NR waveform. The inaccuracy in obtaining the fitting coefficients
means misrepresentation of the NR waveform, which can translate to systematic
error in downstream data analysis task. For example, sufficiently large
systematic errors in the waveform would cause the recovered source parameters to be biased
in the case of PE.

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then
``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some addition
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.
The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
of parameters instead of all together. This means the tuning process ignore the
correlation between different sub-sets of parameters, so the best-fitted
solution obtained by fitting sub-sets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc}

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform.
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}(\lambda)}),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
% Add description of the optimization procedure.

@@

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-off in
performance in different regions of the intrinsic parameter space.

@@

In Fig. \ref{fig:mismatch_hist}, we show the distribution of log mismatches.
One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
showing the optimization improves the accuracy of the waveform model.
% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
While many waveforms improved their accuracy, some remain to have similar mismatches. This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
Nevertheless, the peak of the original waveform distribution moved to the left end of the distribution, indicating that the optimization procedure improved the model. 
Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses.

>>>>

Here, we illustrate how three core tasks in GW science can be substantially improved through the use of differentiable waveforms.
In the paper we primarily look at toy examples, leaving a more careful analyses to future work. 
The three tasks discussed here cover a wide range of GW science, starting with waveform development all the way to Fisher forecasting and PE.

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}.

%\sout{Every waveform approximant has some}
\AZ{Waveform approximants generally have free coefficients which are calibrated
to NR waveforms to achieve high accuracy.}
\sout{ fitting coefficients that are tuned during
the process of calibrating the ansatz to NR waveforms.} In the case of
IMRPhenomD, there are 209 fitting coefficients used to capture the separate
behavior of the amplitude and phase as a function of the mass ratio and spins.
\sout{The accuracy of fitting coefficients determines how well the approximant
reproduces the NR waveform. 
The}
\AZ{Any} inaccuracy in obtaining the fitting coefficients
means misrepresentation of the NR waveform, which can translate to systematic
error in downstream data analysis task. For example, sufficiently large
systematic errors in the waveform would cause the recovered source parameters to be biased
in the case of PE.

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then
``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some addition
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.
The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
of parameters instead of all together. This means the tuning process ignore the
correlation between different sub-sets of parameters, so the best-fitted
solution obtained by fitting sub-sets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted for during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

\AZ{Maybe comment on optimization method used for EOB waveforms?
I think they do interesting Monte Carlo etc. Check what method is used 
for IMRPhenomXPHM?}

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform.
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
% Add description of the optimization procedure.

@@

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-offs in
performance in different regions of the intrinsic parameter space.

@@

In Fig.~\ref{fig:mismatch_hist}, we show the distribution of log mismatches.
One can see the distribution of mismatch after optimization has smaller mismatch compared to before optimization,
showing the optimization improves the accuracy of the waveform model.
% However, it is not improving for every waveforms. Say why briefly, and say there is another paper coming out.
While many waveforms improved their accuracy, some retain similar mismatches. 
\AZ{I don't understand this claim. The spin degeneracy is present in the physics of the waveform; $\chi_{\rm eff}$ is just one way of parametrizing the spins that tries to reduce the degeneracy.}
This phenomenon is mainly due to the reduced spin approximation used in IMRPhenomD, where spin degeneracy is introduced, hence restricting further improvements. 
In an upcoming paper, we will give a detailed discussion of how such an approximation scheme affects the optimization procedure. 
Nevertheless, the peak of the original waveform distribution has moved to the left end of the distribution, indicating that \sout{the} \AZ{our AD-assisted} optimization procedure \AZ{provides an improved implementation of the model}. 
\sout{Therefore, the new set of waveform coefficients can produce better waveforms for GW analyses. }
